Developmental trajectories of word recognition: a priming study in bilinguals

### 2023-11-09 manuscript/manuscript.qmd
Add new comments by NSG

library(knitr)
library(kableExtra)
library(ggplot2)
library(gt)
library(gtExtras)
library(patchwork)
library(english)
library(gtExtras)

@@

Building a mental lexicon is a major achievement in the development of an infant. By storing representations of how familiar words sound and what they mean, an infant is able to make sense of their linguistic input. Infants start forming their first lexical representations before their first year of life [@tincoff1999some; @halle1994emergence; @vihman2004cross; @parise2012electrophysiological; @bergelson20126; @bergelson2015early]. This initial lexicon consists of only a few items; mainly words for people, interjections, body parts, and food [@tincoff2012six; @tardif2008baby], but it undergoes rapid growth during their second year of life [@ganger2004reexamining; @goldfield1990early; @bloom2002children; @mcmurray2007defusing]. According to parental reports the average 15-month-old infant already understands more than 100 words, and by two years, they understand more than 400 words [@frank2021variability]. This accelerated lexical developmental is reflected in infants' trajectories of word recognition: infants recognise familiar words faster and more efficiently as they approach their second birthday [@fernald1998rapid; @fernald2001half; @hurtado2007spoken].

Despite being exposed to a more complex linguistic input, bilinguals show equivalent trajectories of word acquisition and word recognition to their monolingual peers' [@vihman2007onset; @legacy2018vocabulary; @hoff2012dual; @de2014bilingual; @pearson1994patterns; @byers-heinlein2023sometimes]. This is a remarkable deed for two reasons. First, bilingual infants receive a relative impoverished linguistic input in each of their languages compared to monolinguals [@costa2014does]. Second, they face a more complex referential context, since they often learn two labels for each referent, one in each language. The mechanisms that allow bilingual infants to keep up with monolinguals in their lexical development trajectories are still unclear.

Monolinguals and bilinguals share broadly common word-recognition mechanisms. The first step towards spoken word-recognition is the acoustic-phonetic processing of the speech signal, which results in the activation of lexical representations whose associated phonological forms match the acoustic signal. The activated word-forms then compete for selection: the lexical system uses the bottom-up information provided by the auditory input, together with top-down context---provided by grammatical, supra-segmental, and semantic constraints---to select the candidate lexical representation with the best match. This sequence of events occurs in a *cascaded* fashion. As the recognition system accumulates information about the unfolding acoustic signal, matching phonological representations are activated [@marslen1987functional; @marslen1978processing; @grosjean1980spoken], and in turn activate their associated semantic representations [@neely1977semantic]. By the time lexical selection occurs, many non-selected phonological and semantic representations have been activated, potentially modulating the dynamics of word recognition.

A considerable amount of evidence for such cascaded account of lexical processing has been provided by studies recording eye movements during word recognition. For example, @allopenna1998tracking designed a word-recognition task in which participants were presented, in each trial, with four objects in the screen. Participants would then listen to a command like "Pick up the beaker; now put it below the diamond", and perform the action. The authors manipulated the phonological relationship between the target object's label (e.g., *casket*), and the label of each of the three distractors. Two of the distractors were phonologically related with the target, sharing onset (e.g., *castle*) or offset (e.g., *basket*) with the target word, respectively. The other distractor was phonologically unrelated to the target (e.g., *nickel*). After hearing the target label, phonological distractors attracted participants' eye fixations more than unrelated distractors. Distractors sharing phonological onset did so at an earlier time window than those sharing phonological offset. This time course of distractor fixations suggests that the auditory presentation of the target label activated no only its (subsequently selected) phonological form in participants' lexicon, but also that of phonologically related word-forms, leading to competition for selection between the word forms. These results converge with cascaded models of auditory word recognition, in which the unfolding acoustic input activates multiple phonological word-forms, which in turn compete for selection, like TRACE [@mcclelland1986trace] and Shortlist [@norris1994shortlist].

The developing mental lexicon also seems to operate in this cascaded fashion. Like adults, infants activate non-selected lexical representations. This is reflected by the fact that from 21 months of age, infants' lexicon seems to also be systematically organised according to the semantic and phonological properties of its lexical representations. Priming paradigms of word recognition have provided a strong body of evidence for the existence of phonological and semantic links between early lexical representations between 18 and 21 months of age [@arias2009lexical; @styles2009infants]. For instance, using a word recognition task, @arias2009lexical found that infants' spoken word recognition (e.g., *dot*) was interfered by the previous presentation of a semantically related word (e.g., *cat*). Other studies have provided electrophysiological evidence of semantic priming at earlier ages [e.g., @rama2013development; @friedrich2005lexical]. More recent studies have provided evidence for the emergence of such semantic links, even in the absence of visual referents during the experimental task [@willits2013toddlers].

Phonological priming effects have also been reported in the initial lexicon. @mani2010infant created an implicit naming task in which each trial started with the *silent* presentation of a prime picture. Then, a target-distractor picture pair was presented side-by-side. Finally, the target picture's label was presented auditorily. The authors manipulated the phonological overlap between the prime label and the target label, so that in half of the trials both labels were phonologically related, sharing phonological onset (*cat*-*cup*), or phonologically unrelated (*ball*-*comb*). Prime and target-distractor pairs were semantically unrelated. Interestingly, infants showed a stronger looking preference for the target picture after phonologically related prime pictures, compared to after phonologically unrelated primes. This suggests that infants implicitly named prime pictures despite such pictures having been presented in silence, and that the phonology of the resulting label interacted with the subsequent auditory recognition of the target label.The authors found a facilitation effect: target looking preference was stronger in phonologically related primes, indicating that the activated phonological segments of the generated prime label facilitated the subsequent activation of the phonological representation of the target label during recognition.

In a follow-up study, @mani2011phonological found the opposite effect in older toddlers. After phonologically related primes, 24 month-old toddlers showed weaker target looking preference than after phonologically unrelated primes. These results interpreted these outcome as the result of the implicitly generated prime label interfering with the subsequent recognition of the (phonologically related) target word. The authors suggested that the shift from facilitation to interference from 18 to 24 months they found in both studies might be the result of a developmental shift in which 18-month-olds' lexicon is not yet organised based on phonological similarity, while 24-month-olds' is. This would lead to the former showing a pre-lexical facilitation effect in which such facilitation occurs via the activation of phonological segments, while the latter would show an interference effect due to lateral links between lexical representation having been established.

@chow2017spoken adapted the Visual World Paradigm from @huettig2007tug to explore 24- to 30-month-old toddlers' visual fixation patterns during a word recognition when presented with phonological and semantic distractors. In each trial, the authors presented participants with four semantically and phonologically pictures. Four seconds after pictures onset, a word-form was auditorily presented. The word did not refer to any of the pictures displayed on the screen, but was phonologically related to one of them (both labels shared phonological onset), and semantically related to another one of the pictures (both referents belonged to the same taxonomic category. For instance, participants might be presented with the pictures of a sandwich, a bus, a cat, and a dress. Then they would hear the carrier phrase "Look at the *bee*!". The authors registered participants fixations to the phonological and semantic distractors, and found evidence of a preference for the phonological distractor at earlier stages of the post-naming phase, and a preference for the semantic distractor at later stages of the trial. These results support a cascaded activation account of lexical access during the first stages of lexical development, paralleling previous findings in adults [@allopenna1998tracking].

One of the most interesting pieces of evidence in support of a cascaded account of lexical processing is provided by bilinguals. A critical property the bilingual lexicon is that its representations are accessed in a language-non selective way. Even during monolingual situations, bilinguals access their lexical representations in both of their languages [e.g., @thierry2007brain; @dunabeitia2009masked; @HOSHINO; @STROJKERS2010; @dijktra2010baayen]. This is reflected in bilinguals' performance in word recognition and production tasks. For instance, @marian1999activation presented Russian-English bilinguals with a series of trials that started with an auditory instruction like "Poloji marku nije krestika" ["Put the stamp below the cross"]. The instruction would be present in Russian (dominant language for most participants). The target object (*marku* [*stamp*], in this case) would be present in the screen, together with an object whose English label shared phonological onset with the target label (e.g., *mark*), and two unrelated distractors whose labels in Russian and English did not shared phonological relationship with the target label. Converging with @allopenna1998tracking's results in monolingual adults, after hearing the target label bilinguals fixated the cross-language phonological distractor significantly more than the unrelated distractors. These findings suggest that participants activated phonologically related word-forms in both Russian *and* English, which affected their overt visual exploration patterns during the task.

Language non-selectivity in the bilingual lexicon has also been shown to affect top-down lexical access during word production. @costa2000cognate presented Catalan-Spanish bilinguals with a series of pictures of familiar objects. For each object, participants were asked to name it in Catalan (their dominant language). Unbeknownst to participants, the authors manipulated the phonological overlap between the pictures' labels in Catalan and their translations in Spanish. In half of the trials, the labels associated to pictures were cognates, that is, phonologically similar translation equivalents in Catalan and Spanish, whereas in the other half of the trials the labels were non-cognates. Participants produced faster words in cognate trials (e.g., *gat*-*gat* [*cat*]) than in non-cognate trials (e.g., *taula*-*mesa* [*table*]). Critically, Spanish monolinguals who completed the same task, naming the pictures in Spanish, showed equivalent naming times in both conditions. These results reveal that bilinguals activated their Spanish phonology, despite performing the naming task exclusively in Catalan: the visual recognition of the presented pictures led to the activation of its associated phonological forms in both language sin parallel, which influenced the subsequent the dynamics of word production. the differential processing of cognates has generated a large body of literature providing support for the language-non selective account of lexical access [e.g., @degrootandnas1992; @schwartzkroll2007; @dijkstragrainger1999; @dufourkroll1995; @lemhofferetal2008]

Further support for the parallel retrieval of phonological forms during word production is given by implicit naming paradigms. Using an adaptation of @mani2010infant's task, @von2014bilinguals provided evidence that bilingual adults generate implicit labels in both labels for visually fixated pictures. The authors presented German-English bilinguals with 120 prime-target pairs. Primes were presented as familiar pictures in silence. After prime picture offset, target words were presented auditorily. Participants' N400 ERP components were recorded from target word onset. The authors manipulated the phonological relationship between the prime and target word-forms within participants' L1 and across L1 and L2. When prime and target were phonologically identical (*Affe*-*Affe*, German for *monkey*), or similar (*Fahne*-*Sahne*, German for *flag* and *ice-cream*, respectively) within L1, participants' showed a reduced N400 amplitude, compared to when prime and target were phonologically unrelated (*Messer*-*Seil*, German for *knife* and *rope*). Critically, a similar effect was found when prime and target were phonologically related through translation (*Rustsche*-*Kleid*, German for *slide* and *dress*). This suggests that participants activated prime labels in both languages in parallel, and that both labels impacted the dynamics of target words recognition.

Phonological priming paradigms have also revealed language non-selectivity in the initial bilingual lexicon [@poarch2012cross; @bosma2020cognate; @singh2014one; @jardak2019labels; @von2019impact]. For instance, @von2012language found evidence of cross-language phonological priming in 21- to 42-months-old children learning German and English. At the beginning of each trial, the authors auditorily presented an English prime word embedded in a carrier phrase. Then, the target label was also auditorily presented in German. Finally the target and distractor pictures were presented side-by-side. The authors manipulated the phonological overlap between the prime and target labels. The novelty in this study lied in the orthogonal manipulation of the phonological overlap between both English and German labels. In some trials, the auditorily presented prime and target labels were phonologically related (e.g., *slide*-*Klide* [*dress*]). In some other trials, both labels were phonologically related through translation: the auditorily presented prime label did *not* overlap with the target label (e.g., *leg*-*Stein* [*stone*]), but its translation in German did [*Bein*]. In the rest of the trials prime and target labels were phonologically unrelated in both languages. The authors found a facilitation effect of cross-language priming when both prime and target auditory labels overlapped phonologically, as revealed by a stronger target picture looking preference, as compared to that of unrelated trials. Interestingly, participants showed a weaker target preference in priming through translation trials.

Using a similar task, @floccia2020translation provided conflicting results to those of @von2012language. The authors tested 27-month-old simultaneous bilinguals in a cross-language priming paradigm. The two languages participants were learning were English, and an Additional Language (either Cantonese, Dutch, French, German, Greek, Italian, Mandarin, Polish, Portuguese, or Spanish). A prime word was embedded at the end of a carrier sentence that participants listened to. Then, participants were presented with the auditory label of the target word, and two pictures were shown side-by-side, the target picture, and a distractor picture. In Study 1, the target word was the translation equivalent of the prime word (*cheese*-*fromage* for an English-French bilingual). In other trials, prime and target words were unrelated (e.g., *sock*-*fromage*). There was no phonological overlap at onset between prime and target or distractor labels in either of the two conditions. Participants were presented with two blocks of related and unrelated trials. In one of the blocks, participants were presented with English prime words and Home Language target words (*sock*-*fromage*), and in the other block they were presented with Additional Language prime words and English target words (*fromage*-*sock*). The authors found a cross-linguistic priming effect, in which participants showed a stronger preference for the target picture in the post-naming phase in related trials, as compared to unrelated trials. This effect was found regardless of the language of the prime (English or Home Language), or the dominance of the language of the prime (i.e., whether the prime word belonged to the dominant language). Target preference exceeded chance level in both conditions.

In Study 2, the authors tested a similar group of bilinguals in a cross-linguistic semantic priming task, in which participants were primed with words in English, and tested with semantically related words in the Additional Language (or vice versa). This time, prime and target words were not translation equivalents, but rather words whose referents shared semantic features. Results from Study  2 were parallel to those from Study 1. The authors found a strong semantic priming effect across languages, regardless of the language of the prime or the dominance of the language of the prime. Overall these results suggest that bilinguals recognised target words in both conditions, and benefited from a cross-linguistic facilitation effect from translation equivalents. The symmetry of such effect across both languages suggests that participants lexical access occurred in parallel for the two members of translation equivalents. As @floccia2020translation note, the stimuli set of Study 2 contains multiple cognates, especially in languages with a larger degree of phonological overlap across translation pairs, like English and Dutch. The authors run *post hoc* analyses exploring the effect of cognateness on the cross-language priming effects found, which showed no evidence of a cognateness effect on target preference. However, the cross-language effects reported in Study 2 might still reflect the confounding influence of cognateness, as the resulting effect might have been generated by an interaction by the phonological and semantic overlap between cognate primes and targets.

Previous studies had reported similar cross-language priming effects, but as @von2012language note, the conclusion that such effects occurred at the lexical level, as opposed to being the result of mere acoustic similarity, was not warranted. In studies where the auditorily presented prime label was phonologically overlapped with the target label [@REFERENCE], priming effects might be explained by the activation of the phonological segments of the prime label, which facilitated the recognition of the target label, which shared such phonological segments. The priming-through-translation results by @von2012language and @floccia2020translation demonstrates that cross-language activation in the developing lexicon occurs at the lexical level. Since participants' target recognition was interfered by the non-presented prime translation, such an inhibitory effect must stem at the lexical level. The fact that phonological priming facilitated the recognition of the target words (both labels belong to different languages), and that phonological priming through translation interfered target recognition (translation and target belong to the same language) points to cross-language connections playing an excitatory role, and to within-language connections playing an inhibitory role.

Although priming-through-translation studies suggest parallel activation in participants' lexicon in both languages, their design prevents strong interpretations about the extent to which parallel activation is the result of the experimental manipulations of interest. The presentation of primes and targets in both languages in the same task---or even within the same trial--- presents participants with a bilingual context that might lead to a higher baseline activation of lexical representations of both languages, compared to monolinguals situations in which parallel activation is claimed to spontaneously take place [@GROSJEAN1997]. Another limitation of priming-through-translation studies is that, when prime words are auditorily presented, the task constrains the naming context to one of the languages. The consequence is a potential asymmetry between the activation dynamics of the lexical representation of both labels: while the participant has heard one label, the activation of its translation must spread across semantic links. This might lead to a lower strength, and higher latency of the activation of the translation.

In the present study, we aim to put the hypothesis that bilingual infants activate both languages in parallel to a harder test. We use an extension of the implicit naming paradigm by @mani2010infant, in which participants are primed with a picture presented in silence. This provides participants with an exclusively monolingual context in which they perform the task in only one of their languages. If differences in task performance are found associated with covert manipulation of cross-linguistic properties of the stimuli, such differences may be more conclusively linked to the parallel activation of lexical representations in participants' lexicons.

Another advantage of the implicit naming paradigm in the investigation of bilingual lexical processing is the fact that, since participants are not presented with any auditory prime label, the activation of the lexical representation of the prime may occur in both languages as the result of lexical retrieval after visual recognition of the prime picture. This contrasts with priming through-translation tasks, in which the prime label in one of the languages is directly activated by auditory word recognition, while the label of its translation is expected to be activated as the result of cascaded activation spreading through their semantic association. This may lead to an asymmetry in the time course of the activation of both labels, which might in turn affect the dynamics of phonological priming in subsequent word recognition.

Finally, testing bilingual populations at early ages presents the limitation that the vocabulary size of participants' non-dominant language is relatively small, which substantially constrains the number of items that can be used to as stimuli in the task, as well as the number of trials that can be constructed to generate stimuli lists. In order to circumvent the problem of limited vocabulary knowledge in the non-dominant language, tested participants only in their dominant language, therefore directing our research to the question of how possessing a second language affects processing of the dominant one in bilingual toddlers [@costa2014does provide a fuller description of the advantages of such approach].

In summary, the present study was aimed at investigating the developmental trajectories of word recognition in bilingualism, and how (if at all) language non-selectivity shapes them. Using an adaptation of [@mani2010infant], we presented 20- to 32-months-old toddlers with a cross-linguistic phonological priming paradigm. We manipulated both the phonological overlap between the prime word and target word, and the cognate status of the prime word, to test the hypothesis that bilingual toddlers implicitly name familiar pictures in both languages. If so, phonological priming effect should be larger when the prime label in both languages shares phonological overlap with the target words (*Related/Cognate* condition), compared to when only the label of the prime in one language overlaps phonologically with the target (*Related/Non-cognate*). We also also included a control *Unrelated* condition in which regardless of the cognate status of the prime word, neither of the prime labels in any language overlapped phonologically with the target word.

To test cross-language priming effects, we registered participants' time course of target picture looking in the three conditions. We predicted within-language priming effects, revealed as differences in target looking between *Unrelated* and *Related/Non-cognate* trials, as previously reported [@von2012, @MANI2001, and @MANI2011]. We further predicted cross-language priming effects, revealed as differences in target looking between *Related/Non-cognate* and *Related/Cognate* trials. 
This effect should be stronger in bilinguals, if not completely absent in monolinguals, who should be indifferent to the phonological properties of the stimuli in the other language. 
Following previous studies associating the emergence of phonological links between lexical items to the size of children's vocabulary [@MANI2011], we also compared the explanatory power of two indices of receptive vocabulary size (dominant-language vocabulary size and total vocabulary size) against participants' age. We used parental reports of vocabulary checklist to estimate participants' vocabulary size. We computed vocabulary size in the dominant language (L1 vocabulary) as the proportion of words in the dominant language reported by their caregivers to be understood by the child in a vocabulary checklist, and the total vocabulary size as the proportion of words reported as understood by the child in *both* languages. If within-language or cross-language phonological associations in the lexicon emerge as a function of vocabulary growth, the size of the differences between conditions should larger in participants with larger vocabulary sizes. A model including participants' L1 or total vocabulary size as a predictor should then show better performance than a model that instead includes participants age. In the case of bilinguals, if the emergence of cross-language priming effects is associated with the growth of their vocabulary is *both* languages, a model including total vocabulary size as a predictor should perform better than a model that includes L1 vocabulary size. The opposite pattern would indicate that cross-language priming effects emerge as a function of the growth of L1 vocabulary.

@@

n_sessions_lp <- participants_bcn |> 
	count(lp) |> 
	group_split(lp) |> 
	set_names(c("monolingual", "bilingual")) |> 
	map("n")
```

>>>>

library(knitr)
library(kableExtra)
library(ggplot2)
library(gt)
library(gtExtras)
library(patchwork)
library(english)
library(gtExtras)
library(marginaleffects)

@@

Building a mental lexicon is a major achievement in the development of an infant: by storing representations of how familiar words sound and what they mean, an infant is able to make sense of their linguistic input. The foundations of an initial lexicon are in place before their first year of life [@tincoff1999some; @halle1994emergence; @vihman2004cross; @parise2012electrophysiological; @bergelson20126; @bergelson2015early]. This initial lexicon consists of only a few items; mainly words for people, interjections, body parts, and food [@tincoff2012six; @tardif2008baby], but it undergoes rapid growth during the second year of life [@ganger2004reexamining; @goldfield1990early; @bloom2002children; @mcmurray2007defusing]. According to parental reports, the average 15-month-old infant already understands more than 100 words, and by two years, they understand more than 400 words [@frank2021variability]. This accelerated lexical developmental is reflected in infants' trajectories of word recognition: infants recognise familiar words faster and more efficiently as they approach their second birthday [@fernald1998rapid; @fernald2001half; @hurtado2007spoken]. Despite being exposed to a more complex linguistic input, bilinguals show equivalent trajectories of word acquisition and word recognition to their monolingual peers' [@vihman2007onset; @legacy2018vocabulary; @hoff2012dual; @de2014bilingual; @pearson1994patterns; @byers-heinlein2023sometimes]. This is a remarkable deed for two reasons. First, bilingual infants receive a relative impoverished linguistic input in each of their languages compared to monolinguals [@costa2014does]. Second, they face a more complex referential context, since they often learn two labels for each referent, one in each language. The mechanisms that allow bilingual' trajectories of lexical developmental to keep up with monolinguals' are still unclear.

Previous studies have pointed to the interplay between bilingual' languages as a candidate facilitative mechanism behind bilingual lexical acquisition. In particular, cross-language similarity at the lexical level has been associated with larger vocabulary sizes. @floccia2018introduction collected vocabulary data from a sample of 367 bilingual children living in the United Kingdom, who were learning English and an additional language (out of a diverse pool of 13 languages). Using an edit distance-based metric of phonological similarity, the authors computed the average lexical similarity between the translation pairs of the English vocabulary checklist, and of each of the other languages. Children's vocabulary sizes in the additional language was positively associated with the amount of language similarity between their two languages. For instance, children learning English and Dutch (which share high lexical similarity) showed larger vocabulary sizes than those learning English and Mandarin (which share low lexical similarity). The authors suggested that the acquisition of words in the additional language might be facilitated by their cognate status (i.e., being phonologically similar to their translation equivalent). Larger vocabulary sizes might then be expected in bilinguals learning two languages sharing a high proportion of cognates. Recent studies have reported an earlier acquisition of cognate words [@mitchell2022cognates; @garcia-castro2023cognate], providing further evidence that cross-language similarity at the lexical level might boost vocabulary growth in bilinguals.

The facilitative effect of cross-language similarity is in line with the language non-selective account of bilingual lexical access. There is robust evidence suggesting that bilingual adults activate both languages in parallel---even during monolingual situations---which influences their performance in word recognition and production tasks [@marian1999; @groot1992determinants; @schwartz2007reading; @dijkstra1999recognition; @dufour1995matching; @schwartz2007reading; @dijkstra2010cross]. For instance @costa2000cognate presented Catalan-Spanish bilinguals with a series of pictures of familiar objects. For each object, participants were asked to name it in Catalan (their dominant language). Unbeknownst to participants, the authors manipulated the cognate status pictures' labels in Catalan and Spanish. In half of the trials, the labels associated to the pictures were cognates (i.e., phonologically similar translation equivalents, e.g., *gat*-*gat* [*cat*]) in Catalan and Spanish, whereas in the other half of the trials the labels were non-cognates (e.g., *taula*-*mesa* [*table*]). Participants named pictures faster in cognate trials than in non-cognate trials. Critically, Spanish monolinguals who completed the same task, naming the pictures in Spanish, showed equivalent naming times in both conditions. These results revealed that bilinguals activated their Spanish phonology, despite performing the naming task exclusively in Catalan: the visual recognition of the presented pictures led to the activation of its associated phonological forms in both languages in parallel, which influenced the subsequent the dynamics of word production.

Parallel activation has also been reported in he developing lexicon [@poarch2012cross; @bosma2020cognate; @singh2014one; @jardak2019labels; @von2019impact; @floccia2020translation]. For instance, @von2012language found evidence of cross-language phonological priming in twenty 21- to 42-months-old children learning German and English. In their experimental task, each trials begun with the auditory presentation of an English prime word embedded in a carrier phrase. Then, the target label was auditorily presented in German. Finally, the target and a distractor pictures were presented side-by-side. The authors created three conditions. In the phonological priming condition, the auditorily presented prime and target labels were phonologically related (e.g., *slide*-*Klide* [*dress*]). In the priming through translation condition, both labels were phonologically related through translation: the auditorily presented prime label did *not* overlap with the target label (e.g., *leg*-*Stein* [*stone*]), but its translation in German did [*Bein*]. In the unrelated condition, prime and target labels were phonologically unrelated in both languages. Interestingly, participants showed a weaker target preference in priming through translation trials, compared to unrelated trials. This was interpreted evidence of the lexical representations of the prime labels being activated in both English and German, leading to interference effects in later target recognition.

Previous studies had reported similar cross-language priming effects. The conclusion that such effects occurred at the lexical level, as opposed to being the result of mere acoustic similarity, was nonetheless unwarranted. In studies where the auditorily presented prime label was phonologically overlapped with the target label [e.g., @marian1999activation], priming effects might be explained by the activation of the phonological segments of the prime label, which facilitated the recognition of the target label, which shared such phonological segments. This might have been the phonological priming trials in @von2012language, in which prime and target labels shared overt phonological overlap. In contrast, the priming through translation results suggest that cross-language activation in the developing lexicon occurs at the lexical level. Since participants' target recognition was modulated by the non-presented prime translation, such an inhibitory effect must stem at the lexical level. The fact that phonological priming facilitated the recognition of the target words (both labels belong to different languages), and that phonological priming through translation interfered target recognition (translation and target belong to the same language) points to cross-language connections playing an excitatory role, and to within-language connections playing an inhibitory role.

Despite extending the available knowledge on the developing bilingual lexicon, the study by @von2012language has some methodological challenges that prevent strong conclusions about language non-selective access. First, participants' received their English input at school, where they heard German and English (between three and four hours each) from a German native speaker. Except for four children who also listened to English at home, participants' only exposure to English was 3-4 hours per day, for the time they attended daycare (an average of 12.35 months). Assuming that infants are exposed to a daily average of 12 total hours of speech, and they attended daycare only during weekdays, their estimated exposure to English throughout the week represented between `r scales::percent((3*5*4*12.35)/(12*7*4*21), accuracy = 0.01)` and `r scales::percent((4*5*4*12.35)/(12*7*4*21), accuracy = 0.01)` for the 21 month-olds (the youngest in the sample), and between `r scales::percent((3*5*4*12.35)/(12*7*4*43))` and `r scales::percent((4*5*4*12.35)/(12*7*4*43))` for the 43-month-olds. Conventionally, children are considered bilingual if exposed to 25% or more of the time to a second language [@byers-heinlein2021multilab; @rocha-hidalgo2023defining]. Participants in @von2012language might therefore not be generalisable to other populations of bilinguals who received a more balanced dual language input. Second, the presence of cognates in the stimuli lists lead to some primes in the phonological priming condition sharing substantial cross-language phonological overlap thought translation with the target words like *fire* /\textipa{faIe}/--*Feuer* /\textipa{fOyeR}/ and *Eier* /\textipa{aIeR}/, or *Moon* /\textipa{mu:n}/--*Mond* /\textipa{mo:nd}/ and *Huhn* /\textipa{hu:n}/. Conversely, some trials in the phonological priming through translation shared within-language overlap with the target, such as *Soup* /\text{su:p}/--*Suppe* /\textipa{zupe}/---*Puppe* /\text{pupE}/ and *Puppe* /\textipa{pupE}/. Given that the number of total trials in both conditions amounted to six trials, a considerable proportion of the trials presented this methodological limitation. Finally, the observed pattern of results was dependent on the inclusion of a single item, *Buch* [*book*, in German]. When this item was included, monolinguals showed a similar interference effect to bilinguals in the priming through translation condition, in which only bilinguals with knowledge of the translation should show an effect. Additionally, given that monolinguals and bilinguals were not directly, it is not possible to conclude that the differences in statistical significance between monolingual and bilingual target looking, compared to chance levels, does not necessarily provide a significant difference between the two groups. As a consequence, the conclusion that the interference effects observed in the priming through translation condition is exclusive to bilinguals (which would suggest that the priming effect were not due to the intrinsic properties of the stimuli) in unwarranted.

Additionally, although the priming through translation paradigm provides a convenient framework to investigate parallel activation in the bilingual lexicon, its design prevents strong interpretations about the extent to which parallel activation is the result of the experimental manipulations of interest. The presentation of primes and targets in both languages in the same task---or even within the same trial---presents participants with a bilingual context that might lead to a higher baseline activation of lexical representations of both languages, contrary to a context in which parallel activation results from genuine cross-language activation [@grosjean1997bilingual]. Another limitation of priming through translation studies is that, when prime words are auditorily presented, the task constrains the naming context to one of the languages. The consequence is a potential asymmetry between the activation dynamics of the lexical representation of both labels: while the participant has heard one label, the activation of its translation must spread across semantic links. This might lead to a lower strength, and higher latency of the activation of the translation. In summary, current evidence on the cross-language interplay in the initial lexicon should be considered carefully.

In the present study, we put the hypothesis that bilingual infants activate both languages in parallel to a more severe test. We used an extension of the implicit naming task by @mani2010infant, in which participants were primed with a picture presented in silence. Previous studies using this paradigm have successfully provided evidence of language non-selectivity. In particular, @von2014bilinguals found that bilingual adults generate implicit labels in both labels for visually fixated pictures. The authors presented German-English bilinguals with 120 prime-target pairs. Primes were presented as familiar pictures in silence. After prime picture offset, target words were presented auditorily. Participants' N400 ERP components were recorded from target word onset. The authors manipulated the phonological relationship between the prime and target word-forms within participants' L1 and across L1 and L2. When prime and target were phonologically identical (*Affe*-*Affe*, German for *monkey*), or similar (*Fahne*-*Sahne*, German for *flag* and *ice-cream*, respectively) within L1, participants' showed a reduced N400 amplitude, compared to when prime and target were phonologically unrelated (*Messer*-*Seil*, German for *knife* and *rope*). Critically, a similar effect was found when prime and target were phonologically related through translation (*Rustsche*-*Kleid*, German for *slide* and *dress*). This suggests that participants activated prime labels in both languages in parallel, and that both labels impacted the dynamics of target words recognition.

This paradigm provides participants with an exclusively monolingual context in which they perform the task in only one of their languages. If differences in task performance are found associated with covert manipulation of cross-linguistic properties of the stimuli, such differences may be more conclusively linked to the parallel activation of lexical representations in participants' lexicons. By using an implicit naming paradigm, the activation of the lexical representation of the prime may occur in both languages as the result of lexical retrieval after visual recognition of the prime picture, since participants are not presented with any auditory prime label. This contrasts with priming through-translation tasks, in which the prime label in one of the languages is directly activated by auditory word recognition, while the label of its translation is expected to be activated as the result of cascaded activation spreading through their semantic association. This may lead to an asymmetry in the time course of the activation of both labels, which might in turn affect the dynamics of phonological priming in subsequent word recognition.

In the present investigation, we presented 20- to 32-months-old toddlers with a cross-linguistic phonological priming paradigm. We manipulated both the phonological overlap between the prime word and target word, and the cognate status of the prime word, to test the hypothesis that bilingual toddlers implicitly name familiar pictures in both languages. If so, phonological priming effect should be larger when the prime label in both languages shares phonological overlap with the target words (*Related/Cognate* condition), compared to when only the label of the prime in one language overlaps phonologically with the target (*Related/Non-cognate*). We also also included a control *Unrelated* condition in which regardless of the cognate status of the prime word, neither of the prime labels in any language overlapped phonologically with the target word. To test cross-language priming effects, we registered participants' time course of target picture looking in the three conditions. We predicted within-language priming effects, revealed as differences in target looking between *Unrelated* and *Related/Non-cognate* trials, as previously reported [@von2012language, @mani2010infant, and @mani2011phonological]. We further predicted cross-language priming effects, revealed as differences in target looking between *Related/Non-cognate* and *Related/Cognate* trials. 
This effect should be stronger in bilinguals, if not completely absent in monolinguals, who should be indifferent to the phonological properties of the stimuli in the other language.

Following previous studies associating the emergence of phonological links between lexical items to the size of children's vocabulary [@mani2011phonological], we also compared the explanatory power of two indices of receptive vocabulary size (dominant-language vocabulary size and total vocabulary size) against participants' age. We used parental reports of vocabulary checklist to estimate participants' vocabulary size. We computed vocabulary size in the dominant language (L1 vocabulary) as the proportion of words in the dominant language reported by their caregivers to be understood by the child in a vocabulary checklist, and the total vocabulary size as the proportion of words reported as understood by the child in *both* languages. If within-language or cross-language phonological associations in the lexicon emerge as a function of vocabulary growth, the size of the differences between conditions should larger in participants with larger vocabulary sizes. A model including participants' L1 or total vocabulary size as a predictor should then show better performance than a model that instead includes participants age. In the case of bilinguals, if the emergence of cross-language priming effects is associated with the growth of their vocabulary is *both* languages, a model including total vocabulary size as a predictor should perform better than a model that includes L1 vocabulary size. The opposite pattern would indicate that cross-language priming effects emerge as a function of the growth of L1 vocabulary.

Testing bilingual populations at early ages presents the limitation that the vocabulary size of participants' non-dominant language is relatively small, which substantially constrains the number of items that can be used to as stimuli in the task, as well as the number of trials that can be constructed to generate stimuli lists. In order to circumvent the problem of limited vocabulary knowledge in the non-dominant language, tested participants only in their dominant language, therefore directing our research to the question of how possessing a second language affects processing of the dominant one in bilingual toddlers [@costa2014does provide a fuller description of the advantages of such approach].

@@

n_sessions_lp <- participants_bcn |> 
	count(lp) |> 
	group_split(lp) |> 
	set_names(c("monolingual", "bilingual")) |> 
	map("n")

suppressWarnings({
	time_diffs <- participants_bcn |>
		arrange(child_id, session_id, age) |>
		mutate(n_session = 1:n()) |>
		select(child_id, session_id, session_n, age) |> 
		filter(child_id %in% participants_bcn$child_id[duplicated(participants_bcn$child_id)]) |>
		pivot_wider(id_cols = child_id, 
					names_from = session_n, 
					values_from = age, 
					names_repair = janitor::make_clean_names) |> 
		rowwise() |>
		mutate(diff_1 = x2-x1, diff_2 = x3-x2, 
			   diff_min = min(diff_1, diff_2, na.rm = TRUE),
			   diff_max = max(diff_1, diff_2, na.rm = TRUE)) |> 
		ungroup() |> 
		summarise(diff_min = min(diff_min, na.rm = TRUE), 
				  diff_max = max(diff_max, na.rm = TRUE))
})
```

We collected data from `r n_participants_total` children living in the Metropolitan Area of Barcelona (Spain), tested at the Laboratori de Recerca en Infància at the Universitat Pompeu Fabra. Families were recruited from maternity rooms in private hospitals and social media, and contacted via phone when the child's age spanned between 20 and 32 months. From the `r n_participants_total` children that participated, `r n_participants_sessions$session_1` participated once, `r n_participants_sessions$session_2` participated twice, and `r n_participants_sessions$session_3` participated three times. Recurrent participants were tested with at least `r round(time_diffs$diff_min[1], 2)` months of difference. We gathered a total of `r n_sessions_total$n` testing sessions. Participants were divided into monolinguals and bilinguals based on their relative degree of exposure to Catalan and Spanish, estimated using the Language Exposure Questionnaire [LEQ, @bosch2001evidence]. We categorised participants as monolingual if exposed to more than 80% or more of the time to their dominant language, and as bilingual otherwise. `r english::Words(n_participants_lp$n[1])` of the participants were categorised as monolinguals (`r n_participants_lp_sex$n[1]` female, `r n_participants_lp_sex$n[2]` male) and `r n_participants_lp$n[2]` as Catalan/Spanish bilinguals (`r n_participants_lp_sex$n[3]` female, `r n_participants_lp_sex$n[4]` male) (see @tbl-participants for a detailed summary of participants' age and language profile). Participants' vision was normal, none used glasses or any other type of vision corrector.
